A sheaf--function dictionary for algebraic tori over local fields

### 2013-08-17 shanghai.tex
Add ~ to \refs

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves.
Since it is a rigid symmetric monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the sheaf-function dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}, we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show that it is an isomorphism in Theorem \ref{thm:snake}, assuming only
that the geometric component group of $G$ is finitely generated.

In Part 2 we use Theorem~\ref{thm:snake} to make a sheaf--function dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fiber $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $d$, let $\GN{T}_d$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^d}$; the commutative
group scheme $\GN{T}_d$ plays the r\^{o}le of $G$.  Theorem \ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $d$ into quasicharacter sheaves on $\GN{T}_d$:
\[
\QCiso{\GN{T}_d} 
\iso
\Hom_{< d}(T(K),\EEx).
\]  
We refer to finite-depth characters of $T(K)$ as quasicharacters and
describe a group scheme $\GN{T}$ simultaneously geometrizing all such quasicharacters.

We have not yet fully explored the applications of such a geometrization, but want to highlight
some of the potential benefits of Theorem \ref{thm:application}.  Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other reductive groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}, and more recently Drinfeld has referred to such objects
as character sheaves.  Our quasicharacter sheaves evolved from this notion, and Theorem \ref{thm:snake}
is a direct generalization of the sheaf-function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The other main use
of the term ``character sheaf'' traces back to Lusztig \cite{lusztig:85a}.  Most of his energy is devoted to
handling non-commutative groups; though he does define character sheaves for some
non-connected groups he restricts his attention to reductive groups with cyclic component group.
For extensions of $\ZZ/n\ZZ$ by a torus, it is not difficult to relate his Frobenius-stable character
sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section \ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with a Galois action.
Section \ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the sheaf-function dictionary.
We define pullbacks and products of quasicharacter sheaves in Section \ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section \ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections \ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition \ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section \ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give an isomorphism between $\QCiso{G}$ and $\Hh^1(\Weil{\Fq}, \Hom(G(\bFq), \EEx))$.
Section \ref{sec:etalecohom} proves a result in Galois cohomology that shows that trace of Frobenius is
an isomorphism for \'etale group schemes.
In Section \ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section \ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem, that trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

In Section \ref{sec:GN} we introduce N\'eron models of tori and the Greenberg transform, which produce
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section \ref{sec:bdchar} then uses Theorem \ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section \ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections \ref{sec:bdchar} and \ref{sec:quasichar} also include
functoriality results that we hope to use in a future discussion of class field theory.
Finally, we close the paper in Section \ref{sec:transfer} by discussing a transfer principle, which allows
a geometric transfer of characters between characteristic $p$ tori to characteristic $0$ tori.

@@

One of the main results of \cite{chai-yu:01a} gives an isomorphism of group schemes between
$\TT_d$ and $\TT'_d$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{chai-yu:01a}*{\S 8.1}.
Then it follows from \cite{chai-yu:01a}*{Thm. 8.5} that if $N > 3h$ and $T$ and $T'$ are $N$-congruent 
then there is a canonical isomorphism of smooth group schemes
 \[
  \TT_{N-3h} \to \TT'_{N-3h}
 \]
 determined by $\alpha, \beta$ and $\phi$.
Combining this with Theorem \ref{thm:application} gives the geometrization and categorification of the transfer of characters.

>>>>

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves.
Since it is a rigid symmetric monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the sheaf-function dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}, we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show that it is an isomorphism in Theorem~\ref{thm:snake}, assuming only
that the geometric component group of $G$ is finitely generated.

In Part 2 we use Theorem~\ref{thm:snake} to make a sheaf--function dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fiber $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $d$, let $\GN{T}_d$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^d}$; the commutative
group scheme $\GN{T}_d$ plays the r\^{o}le of $G$.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $d$ into quasicharacter sheaves on $\GN{T}_d$:
\[
\QCiso{\GN{T}_d} 
\iso
\Hom_{< d}(T(K),\EEx).
\]  
We refer to finite-depth characters of $T(K)$ as quasicharacters and
describe a group scheme $\GN{T}$ simultaneously geometrizing all such quasicharacters.

We have not yet fully explored the applications of such a geometrization, but want to highlight
some of the potential benefits of Theorem~\ref{thm:application}.  Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other reductive groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}, and more recently Drinfeld has referred to such objects
as character sheaves.  Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf-function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The other main use
of the term ``character sheaf'' traces back to Lusztig \cite{lusztig:85a}.  Most of his energy is devoted to
handling non-commutative groups; though he does define character sheaves for some
non-connected groups he restricts his attention to reductive groups with cyclic component group.
For extensions of $\ZZ/n\ZZ$ by a torus, it is not difficult to relate his Frobenius-stable character
sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with a Galois action.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the sheaf-function dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give an isomorphism between $\QCiso{G}$ and $\Hh^1(\Weil{\Fq}, \Hom(G(\bFq), \EEx))$.
Section~\ref{sec:etalecohom} proves a result in Galois cohomology that shows that trace of Frobenius is
an isomorphism for \'etale group schemes.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem, that trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

In Section~\ref{sec:GN} we introduce N\'eron models of tori and the Greenberg transform, which produce
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses Theorem~\ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar} and \ref{sec:quasichar} also include
functoriality results that we hope to use in a future discussion of class field theory.
Finally, we close the paper in Section~\ref{sec:transfer} by discussing a transfer principle, which allows
a geometric transfer of characters between characteristic $p$ tori to characteristic $0$ tori.

@@

One of the main results of \cite{chai-yu:01a} gives an isomorphism of group schemes between
$\TT_d$ and $\TT'_d$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{chai-yu:01a}*{\S 8.1}.
Then it follows from \cite{chai-yu:01a}*{Thm. 8.5} that if $N > 3h$ and $T$ and $T'$ are $N$-congruent 
then there is a canonical isomorphism of smooth group schemes
 \[
  \TT_{N-3h} \to \TT'_{N-3h}
 \]
 determined by $\alpha, \beta$ and $\phi$.
Combining this with Theorem~\ref{thm:application} gives the geometrization and categorification of the transfer of characters.
